### 2021-07-02 rad-icfp-2021/figures/compilation_scheme.tex
minor

\[
\begin{tikzcd}
    \Source \ar[r] & \UNFSymbol_1 \ar[r,"\Dsynrevsymbol"] & \UNFSymbol_2 \ar[r] & \Target \ar[r,"\partial ev"] & \Target
\end{tikzcd}
\]

>>>>

\[
\begin{tikzcd}
    \Source \ar[rrrr,"\text{efficient }\Dsynrevsymbol (Fig.~\ref{fig:direct_diff_macro})"] \ar[dr,"(Fig.~\ref{fig:source_to_unf})"'] &&&& \Target \arrow[r,loop right,"\text{optim }(Fig.~\ref{fig:optim})"] \\
    & Source\UNFSymbol \ar[rr,"\Dsynrevsymbol (Fig.~\ref{fig:diff_macro})"] && Target\UNFSymbol \ar[ur,"(Fig.~\ref{fig:unf_to_target})"'] & 
\end{tikzcd}
\]

### 2021-07-02 rad-icfp-2021/figures/notation_D.tex
rewrote notation
minor

Ground types are defined inductively by 
(TODO: chose whether we want n-ary, binary, or both kind of products)
$$G::= \RR \mid G\times \ldots \times G \mid A[G]$$

$(\RR,+,\underline{0})$ forms a monoid and this monoid structure extends canonically 
to a monoid structure $(G,\widehat{+},0_G)$ for every ground type $G$. 
It is defined inductively on $G$ as follows
\begin{itemize}
    \item $0_\RR \defeq \underline{0}$
    \item $0_{G_1\times G_2}\defeq < O_{G_1}, 0_{G_2} >$
    \item $0_{A[\RR]_n}\defeq$ ZerosLike($n$) TODO: general array? what about index n?
    \item $a\widehat{+}_\RR b\defeq a+b$
    \item $(a_1,\ldots,a_n)\widehat{+}_{G_1\times\ldots\times G_n}(b_1,\ldots,b_n)\defeq (a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n)$
    \item $A\widehat{+}_{A[\RR]}B \defeq$ !map! + $A$ $B$
\end{itemize}

@@

\begin{notation}
    \begin{itemize}
        \item Given a context $\Gamma=x_1:A_1,\ldots,x_n:A_n$, $Id_\Gamma\defeq$ !fun! $(y_1,\ldots,y_n)$! -> !$(y_1,\ldots,y_n)$
        \item 
        \item If $\Gamma\vdash e:\RR$, $\grad_\Gamma e 
                    \defeq \directD{\Gamma}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,\underline{1})$
        \item If $\Gamma=\Gamma_1,\Gamma_2$ and $|\Gamma_1|=k$ and $\grad_{\Gamma_1,\Gamma_2}e=(d_1,\ldots,d_n)$, 
                then $\grad_{\Gamma_1}(e)\defeq(e_1,\ldots,e_k)$ and $\grad_{\Gamma_2}(e)\defeq(e_{k+1},\ldots,e_n)$
        \item Given a context $\Gamma=x_1:A_1,\ldots,x_n:A_n$ and a variable !x!, we define !pos x! to be its position $i$ in $\Gamma$, assuming all variables are distinct and that $x$ is indeed in $\Gamma$.
        \item Given $G_1,\ldots,G_n$ ground types, $1\leq i\leq n$ and a term !e! of type $G_i$, 
                $[i]e \defeq (0_{G_1},\ldots,0_{G_{i-1}},e,0_{G_{i+1}},\ldots,0_{G_n})$
    \end{itemize}
\end{notation}

>>>>

Ground types are defined inductively by 
$$G::= \RR \mid G\times \ldots \times G \mid A[G]$$

$(\RR,+,\underline{0})$ forms a monoid and this monoid structure extends canonically 
to a monoid structure $(G,\widehat{+},0_G)$ for every ground type $G$. 
It is defined inductively on $G$ as follows
\begin{itemize}
    \item $0_\RR \defeq \underline{0}$
    \item $0_{G_1\times \ldots \times G_n}\defeq < O_{G_1},\ldots, 0_{G_n} >$
    \item $0_{A[\RR]_n}\defeq$ ZerosLike($n$) TODO: general array? what about index n?
    \item $a\widehat{+}_\RR b\defeq a+b$
    \item $(a_1,\ldots,a_n)\widehat{+}_{G_1\times\ldots\times G_n}(b_1,\ldots,b_n)\defeq (a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n)$
    \item $A\widehat{+}_{A[\RR]}B \defeq$ !map! + $A$ $B$
\end{itemize}

@@

\input{figures/notation}

### 2021-07-02 rad-icfp-2021/purerad.tex
minor

TODO: partial evaluation to get a tuple (i.e no lambdas)

Though this transformation already has the right complexity, 
its purity allows us to take the most out of generic optimisations.  
A list of common optimisations that are useful for this language can be found in Figure~\ref{fig:optim}.

TODO: use the optimisations to be efficient on example

>>>>

Following the insight from Section~\ref{subsec:insights}, 
we don't want to keep all the lambda abstractions as this is costly. 
Fortunately, the transformation is designed in such a way that we only have applications 
and we can use partial evaluation to do beta redudction.

By inspection in Figure~\ref{fig:direct_diff_macro}, 
we see that there is a linear usage of  each continuation $Y$ 
and that it is always applied to almost the identity. 
More precisely, we have applications of the form !fun (x1,...,xn) -> Y(e1,...,en)! 
where the !ei! are !xi! except for at most $k$ (independant of $n$) terms.
In fact we have $k=2$, except for the !map2! case which requires a more detailed analysis (see Section~\ref{sec:complexity}).

Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
!fun (x1,...,xn) -> (fun (y1,...,yn) -> (f1,...,fn))(e1,...,en)! 
$\transto$ 
!fun (x1,...,xn) -> let y1,...,yn = e1,...,en in (f1,...,fn)!
$\transto$
!fun (x1,...,xn) -> (f1[e1/y1],...,fn[en/yn])!
Now, importantly, every !yi! is used at most once (except for !map2! again and !reduce! which require a more detailed analysis)
which garantees that this will not blow up the complexity.

TODO: finish explanation, not sure how to conclude at the moment

Though this transformation after the inlining above already has the right complexity, 
its purity allows us to take the most out of generic optimisations.  
A list of common optimisations that are useful for this language can be found in Figure~\ref{fig:optim}.

TODO: use the optimisations to be efficient on examples
